\thetitle

### 2020-11-04 writing/mutate.tex
mutate: Add definition of interesting debugging scenarios

Of course, the gradual typing benchmarks have no mistakes for the rational
programmer to debug. Therefore, we follow
~\citet{lksfd-popl-2020} and we inject bugs with mutation analysis. 
A mutations is a local syntactic change of the code of a program that may
produce a bug. Usually, the outcome of mutating a program
once, dubbed a mutant, can be distinguished from the original program with a test case. 
The better the test suite of a program, the more mutants it can ``kill.''
Thus a big part of research on mutation analysis goes into designing
operations that perform mutations, dubbed mutators, that create mutants 
that are hard to ``kill.'' Unlike the standard mutators, which \citet{lksfd-popl-2020}
use, too, the mutators we need to evaluate the rational programmer have to be
fine-tuned to generate shallow errors. Typed Racket's type system detects 
typos not logical mistakes, and mutators suitable for our purposes
should target the former rather than the latter.

The table in figure~\ref{table:mutation-ops}  summarizes our mutators. 
For each one, the table provides a short description and an example
mutation it performs. Each mutator is inspired by the more-than-a-decade
long experience of the authors making type errors in Typed Racket. 
Our first ten mutators capture usual common in every type language. 
For instance, position swaps the arguments of a function call.d
The last three mutator target distinguishing features of Typed Racket. 
Specifically, \texttt{arithmetic} may replace a \texttt{+} with a \texttt{-} 
in an attempt to change the type of the result of the arithmetic
operation. In Typed Racket, \texttt{+}'s result is a
\texttt{Positive-Integer} when all arguments are
\texttt{Positive-Integer}. However, the result of \texttt{-} is 
\texttt{Integer}. In the same spirit, \texttt{boolean} aims to take
advantage of Typed Racket's ``truthiness.'' Finally,
\texttt{conditional}'s goal is to confuse occurrence typing.

@@

    \texttt{constant}
& Swap a literal constant with another of the same value but different type.
& \texttt{5.6} $\rightarrow$ \texttt{5.6+0.0i} \\
\hline 
    \texttt{deletion}
    & Deletes the result expression of a \texttt{begin} or \texttt{begin0}
    sequence.
    & \texttt{(begin x y z)} $\rightarrow$ \texttt{(begin x y)} \\
\hline

 \texttt{id}
 & Swaps a use of a top-level identifier defined with with another defined
 in the same module.
 &\texttt{(f ...)} $\rightarrow$ \texttt{(g ...)}\\
\hline

    \texttt{list}
    & Swap \texttt{car} and \texttt{cdr}.
    & \texttt{car} $\rightarrow$ \texttt{cdr} \\
 \hline

   \texttt{position}
 & Swap the position of subexpressions.
  &  \begin{tabular}[t]{@{}l}
     \texttt{(f a 42 "b" 0)}\\
       $\rightarrow$\\ 
      \texttt{(f a 42 0 "b")}
    \end{tabular}\\
\hline

\texttt{class:init}
  & Swap default values of class fields.
   &  \begin{tabular}[t]{@{}l}
     \texttt{(class object\%}\\
     \texttt{\,\,(field [a 5] [b "hello"]))}\\
       $\rightarrow$\\ 
      \texttt{(class object\%}\\
     \texttt{\,\,(field [a "hello"] [b 5]))}\\
    \end{tabular}\\
\hline

\texttt{class:method}
 & Add an extra trivial public method to a class.
 &  \begin{tabular}[t]{@{}l}
   \texttt{(class object\%  ...)}\\
       $\rightarrow$\\ 
     \texttt{(class object\%}\\
     \texttt{\,\,(define/public (extra x) x)}\\
     \texttt{\,\,...)}
    \end{tabular}\\
\hline

\texttt{class:parent}
    & Replace the parent of classes with \texttt{object\%}.
    &  \begin{tabular}[t]{@{}l}
     \texttt{(class foo\% ...)}\\
       $\rightarrow$\\ 
      \texttt{(class object\% ...)}\\
    \end{tabular}\\
\hline

\texttt{class:public}
    & Make a public method private and vice versa.
    & 
    \begin{tabular}[t]{@{}l}
      \texttt{(class object\%}\\
        \texttt{\,\,(define/public (m x) x))}\\
       $\rightarrow$\\ 
      \texttt{(class object\%}\\
        \texttt{\,\,(define/private (m x) x))}\\
    \end{tabular}\\
\hline

 \texttt{class:super}
    & Remove \texttt{super-new} calls from class definitions.
    &  \begin{tabular}[t]{@{}l}
      \texttt{(class object\% (super-new))}\\
       $\rightarrow$\\ 
      \texttt{(class object\% (void))}\\
    \end{tabular}\\
\hline

\texttt{arithmetic} 
& Swap an arithmetic operator with another one of the same (or greater)
    arity and a possibly different result type. 
    & \texttt{+} $\rightarrow$ \texttt{-} \\
\hline

\texttt{boolean} 
& Swap \texttt{and} and \texttt{or}. Due to Racket's truthiness it
    possibly changes the type of the operation's result. 
& \texttt{and} $\rightarrow$ \texttt{or} \\

\texttt{conditional}
    & Negate conditional test expressions.
    &  \begin{tabular}[t]{@{}l}
      \texttt{(cond [(= x 0) 42] ...)}\\ 
       $\rightarrow$\\ 
      \texttt{(cond [(not (= x 0)) 42] ...)}
    \end{tabular}\\
\hline

\end{tabular}
  \caption{Summary of mutators.}
  \label{table:mutation-ops}
\end{figure}

An effective set of mutators for type errors should be able to generate
a large number of ill-typed programs. Hence, to validate that our experience 
with Typed Racket translates to effective mutators, we analyze the
 mutants of our ten benchmarks. Figure~\ref{fig:mutators-static-hit} shows
 the results. For each benchmark, our mutators generate between 156 and
 18165 ill-typed mutants. Most importantly the collective success rate of
 the mutators  per benchmark, i.e., the percentage of ill-typed mutants 
 over the number of mutants, ranges between 53\% (\texttt{tetris})
 and 95\% (\texttt{kCFA}). Finally, the figure depicts the individual success rate 
 of mutators per benchmark which reveals that there is at
 least one benchmark where each mutator performs well
 (>90\%).

>>>>

Of course, the gradual typing benchmarks have no mistakes for the rational
programmer to debug. Therefore, we follow
~\citet{lksfd-popl-2020} and we inject bugs with mutation analysis.
A mutation is a local syntactic change of the code of a program that may
produce a bug.
\ll{Need citations for mutation}
Usually, the outcome of mutating a program
once, dubbed a mutant, can be distinguished from the original program with a test case.
The better the test suite of a program, the more mutants it can ``kill.''
Thus a big part of research on mutation analysis goes into designing
operations that perform mutations, dubbed mutators, that create mutants
that are hard to ``kill.'' Unlike the standard mutators, which \citet{lksfd-popl-2020}
use, too, the mutators we need to evaluate the rational programmer have to be
fine-tuned to generate shallow errors. Typed Racket's type system detects
typos not logical mistakes, and mutators suitable for our purposes
should target the former rather than the latter.

The table in figure~\ref{table:mutation-ops} summarizes our mutators.
For each one, the table provides a short description and an example
mutation it performs. Each mutator is inspired by the more-than-a-decade
long experience of the authors making type-level mistakes in Typed Racket.
We find that these mistakes often take non-trivial effort to debug.

Most of our mutators in figure~\ref{table:mutation-ops} are self-explanatory.
The first twelve can apply to most gradually typed languages, including those with classes.
The last four mutators target distinguishing features of Typed Racket.
Specifically, \texttt{arithmetic} may replace a \texttt{+} with a \texttt{-}
in an attempt to change the type of the result of the arithmetic
operation. In Typed Racket, \texttt{+}'s result is a
\texttt{Positive-Integer} when all arguments are
\texttt{Positive-Integer}. However, the result of \texttt{-} is
\texttt{Integer}. In the same spirit, \texttt{boolean} aims to take
advantage of Typed Racket's ``truthiness.'' Finally, \texttt{negate-cond} and \texttt{force-cond}
aim to confuse occurrence typing.

@@

    \texttt{constant}
& Swap a literal constant with another of the same value but different type.
& \texttt{5.6} $\rightarrow$ \texttt{5.6+0.0i} \\
\hline

    \texttt{deletion}
    & Deletes the result expression of a sequence.
    & \texttt{(begin x y z)} $\rightarrow$ \texttt{(begin x y)} \\
\hline

    \texttt{top-level-id}
 & Swaps two identifiers defined in the same module.
 &\texttt{(f x 42)} $\rightarrow$ \texttt{(g x 42)}\\
\hline

    \texttt{imported-id}
 & Swaps two identifiers imported from the same module.
 &\texttt{(f x 42)} $\rightarrow$ \texttt{(g x 42)}\\
\hline

    \texttt{method-id}
 & Swaps two method identifiers.
 &\texttt{(send o f x 42)} $\rightarrow$ \texttt{(send o g x 42)}\\
\hline

    \texttt{field-id}
 & Swaps two field identifiers.
    & \begin{tabular}[t]{@{}l} \texttt{(get-field o f)} $\rightarrow$ \\
        \texttt{(get-field o g)}
      \end{tabular}\\
\hline

   \texttt{position}
 & Swaps two subexpressions.
  &  \texttt{(f a 42 "b" 0)} $\rightarrow$\texttt{(f a 42 0 "b")} \\
\hline

     \texttt{list}
    & Replaces \texttt{append} with \texttt{cons}.
    &  \begin{tabular}[t]{@{}l}
      \texttt{append} $\rightarrow$ \texttt{cons}
    \end{tabular}\\
\hline

     \texttt{class:init}
  & Swaps values of class initializers.
   &  \begin{tabular}[t]{@{}l}
     \texttt{(new c [a 5] [b "hello"])} $\rightarrow$\\
     \texttt{(new c [a "hello"] [b 5])}\\
    \end{tabular}\\
\hline

     \texttt{class:parent}
    & Replaces the parent of classes with \texttt{object\%}.
    &  \begin{tabular}[t]{@{}l}
     \texttt{(class foo\% (super-new))} $\rightarrow$\\
     \texttt{(class object\% (super-new))}\\
    \end{tabular}\\
\hline

     \texttt{class:public}
    & Makes a public method private and vice versa.
    & \begin{tabular}[t]{@{}l}
      \texttt{(class object\%}\\
      \phantom{(d}\texttt{(define/public (m x) x))} $\rightarrow$\\
      \texttt{(class object\%}\\
      \phantom{(d}\texttt{(define/private (m x) x))}\\
    \end{tabular}\\
\hline

     \texttt{class:super}
    & Removes \texttt{super-new} calls from class definitions.
    &  \begin{tabular}[t]{@{}l}
      \texttt{(class foo\% (super-new))} $\rightarrow$\\
      \texttt{(class foo\% (void))}\\
    \end{tabular}\\
\hline

     \texttt{arithmetic}
& Swaps arithmetic operators.
    & \texttt{+} $\rightarrow$ \texttt{-} \\
\hline

     \texttt{boolean}
& Swaps \texttt{and} and \texttt{or}.
& \texttt{and} $\rightarrow$ \texttt{or} \\
\hline

     \texttt{negate-cond}
    & Negates conditional test expressions.
    &  \begin{tabular}[t]{@{}l}
      \texttt{(if (= x 0) t e)} $\rightarrow$\\
      \texttt{(if (not (= x 0)) t e)}
       \end{tabular}\\
\hline

     \texttt{force-cond}
    & Replaces conditional test expressions with \texttt{\#t}.
    &  \begin{tabular}[t]{@{}l}
      \texttt{(if (= x 0) t e)} $\rightarrow$\\
      \texttt{(if \#t t e)}
    \end{tabular}\\
\hline
\end{tabular}
  \caption{Summary of mutators.}
  \label{table:mutation-ops}
\end{figure}

\subsection{Do these mutators give us interesting debugging scenarios?}
\begin{itemize}
\item do these mutators work?
\begin{itemize}
\item now we need to see if these mutators actually produce interesting errors
\begin{itemize}
\item define interesting errors: type error + dynamic err w/ 3 modules on stack in a mixed configuration
\item this section will show if they do this on our benchmarks
\end{itemize}
\item do they actually produce interesting errors? -- for this need plot that's inverted version dyn-err-heatmap + total of mutants per benchmark
\item do we need all of the mutators?
\begin{itemize}
\item each one produces type errors + interesting scenarios
\item we conjecture they represent different kinds of mistakes
\end{itemize}
\end{itemize}
\end{itemize}

Having selected a set of mutators based on our experience, we next need to see if the mutators do in fact produce interesting debugging scenarios in our programs.
We define an interesting debugging scenario as a mutation that
\begin{enumerate}
\item can be caught by the type checker if the program is fully typed,
\item otherwise produces a dynamic error under Erasure, and
\item the stacktrace of the dynamic error contains at least three distinct modules from the program.
\end{enumerate}
The first criterion reflects the fact that a gradual type system can only catch bugs that would have been caught by a type checker.
The second criterion ensures that the bug is detectable by all three semantics.
It is based on~\citet{gfd-oopsla-2019}'s proof that a dynamic type error in Erasure implies a dynamic type error in both Transient and Natural.
The final criterion conservatively approximates scenarios with the opportunity for interactions between typed and untyped modules. % I say opportunity because it depends on the config
In particular, it takes at least two modules to have typed/untyped module interactions (one of each kind), and having two modules on the stack ensures that the error arose in the interaction of at least those two modules.
The third module comes in because every one of our programs comes with a driver module which is usually uninteresting but should be present in every stacktrace.
With this definition in hand, the rest of this section will analyze how well our mutators produce interesting debugging scenarios.

Beginning with the first criterion of the definition, we analyze how well our mutators produce type errors.

%% Cut from above
An effective set of mutators for type errors should be able to generate
a large number of ill-typed programs. Hence, to validate that our experience
with Typed Racket translates to effective mutators, we analyze the
 mutants of our ten benchmarks. Figure~\ref{fig:mutators-static-hit} shows
 the results. For each benchmark, our mutators generate between 156 and
 18165 ill-typed mutants. Most importantly the collective success rate of
 the mutators  per benchmark, i.e., the percentage of ill-typed mutants
 over the number of mutants, ranges between 53\% (\texttt{tetris})
 and 95\% (\texttt{kCFA}). Finally, the figure depicts the individual success rate
 of mutators per benchmark which reveals that there is at
 least one benchmark where each mutator performs well
 (>90\%).
